# Read in and format data
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(brms)
library(ggplot2)

# read in data
eab <- read.table("data/EAB_data_whole.txt", header = T)
gdd <- read.table("data/EAB_GDD_vs_Date.txt", header = T)
dayb <- read.table("data/EAB_covariate_matrix_day_B.txt", header = T)
eab2 <- read.table("data/EAB_data_with_qPCR_hits.txt", header = T)

# format data for GAM fitting
eab.gamdat <- eab2 %>%
  pivot_longer(2:27) %>%
  left_join(gdd, by = c("name"="Day")) %>%
  filter(is.na(value)==F) %>%
  rename(tree = Tree, gdd = GDD, 
         numpos = value) %>%
  mutate(dir = stringr::str_sub(name, -1, -1),
         pos = as.numeric(ifelse(numpos>0,1,0)),
         pos2 = as.numeric(ifelse(numpos>1,1,0))) %>%
  select(-Date.Direction, -name)
```
# Explore the data
```{r}
# count number of visits per tree
visitcount <- apply(eab[,2:27], 1, function(x) sum(is.na(x)==F))
hist(visitcount)
summary(visitcount)

# how many trees have at least 1 hit? 20
  # 2 trees (Tree2 and Tree8) have 0 hits, but only 2 sample days
```
# Run GAMs for detections vs. growing degree days
https://fromthebottomoftheheap.net/2018/04/21/fitting-gams-with-brms/
```{r}
# run a model with only a random intercept for "tree"
tree <- brm(bf(pos ~ (1|tree)),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, 
          control = list(adapt_delta = 0.99),
          save_pars = save_pars(all = TRUE))
saveRDS(tree, "output/tree.gam.rds") # save the output
stancode(tree) # peek at the underlying stan code

# run the same model but now with a smoothed effect of GDD
gdd.tree <- brm(bf(pos ~ s(gdd) + (1|tree)),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, 
          control = list(adapt_delta = 0.99),
          save_pars = save_pars(all = TRUE))
saveRDS(gdd.tree, "output/gdd.tree.gam.rds") # save it
stancode(gdd.tree) # peek at the underlying stan code

# run the same model but now with a categorical effect of "direction"
gdd.tree.dir <- brm(bf(pos ~ s(gdd) + (1|tree) + dir),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, save_pars = save_pars(all = TRUE),
          control = list(adapt_delta = 0.99))
saveRDS(gdd.tree.dir, "output/gdd.tree.dir.gam.rds") # save it
stancode(gdd.tree.dir) # peek at the underlying stan code

# run the same model but with a only categorical effect of "direction"
tree.dir <- brm(bf(pos ~ (1|tree) + dir),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000,
          control = list(adapt_delta = 0.99),
          save_pars = save_pars(all = TRUE))
saveRDS(tree.dir, "output/tree.dir.gam.rds") # save it
stancode(tree.dir) # peek at the underlying stan code
```
# read in the saved models (so we don't have to fit them each time)
And examine the output
```{r}
tree <- readRDS("output/tree.gam.rds")
gdd.tree <- readRDS("output/gdd.tree.gam.rds")
gdd.tree.dir <- readRDS("output/gdd.tree.dir.gam.rds")
tree.dir <- readRDS("output/tree.dir.gam.rds")

# examine the output summary for each
tree
gdd.tree
gdd.tree.dir
tree.dir

# compare the 3 models based on LOOIC
tree <- add_criterion(tree, criterion = "loo",
                          moment_match = TRUE)
gdd.tree <- add_criterion(gdd.tree, criterion = "loo",
                          moment_match = TRUE, reloo = TRUE)
gdd.tree.dir <- add_criterion(gdd.tree.dir, criterion = "loo",
                          moment_match = TRUE)
tree.dir <- add_criterion(tree.dir, criterion = "loo",
                          moment_match = TRUE)

loo_compare(tree, gdd.tree, gdd.tree.dir, tree.dir,
            criterion = "loo") %>% 
  print(simplify = F)

# elpd_diff gdd.tree vs. gdd.tree.dir was -0.8 (SE = 0.3)
# vs. tree.dir it was -14.4 (SE = 5.8)
#   vs. tree it was -13.2 (SE = 5.8)
```
# Plot detection probability vs. GDD smooth
```{r}
# load the model fit
gdd.tree <- readRDS("output/gdd.tree.gam.rds")

# make quick plots to inspect
gdd.tree.smooth <- conditional_smooths(gdd.tree, prob = c(0.95))
gdd.tree.smooth.e <- conditional_effects(gdd.tree, prob = c(0.95))
plot(gdd.tree.smooth)
plot(gdd.tree.smooth.e)

# get predictions vs. GDD with 95 and 80% credible intervals
gdd.tree.smooth.95 <- conditional_effects(gdd.tree, 
                                          prob = c(0.95))$gdd %>%
  select(gdd, p = estimate__, q2.5 = lower__, q97.5 = upper__)

gdd.tree.smooth.80 <- conditional_effects(gdd.tree, 
                                          prob = c(0.80))$gdd %>%
  select(gdd, p = estimate__, q10 = lower__, q90 = upper__)

# get observed proportions on each sampling date
obs <- eab.gamdat %>%
  group_by(as.character(gdd)) %>%
  summarise(obs_p = mean(pos),
            n = length(pos),
            .groups = "drop") %>%
  rename(gdd = 1) %>%
  mutate(gdd = as.numeric(gdd)) %>%
  arrange(gdd) %>%
  mutate(se = sqrt((obs_p * (1-obs_p))/n))

# make the plot
(
gdd.plot <-
gdd.tree.smooth.95 %>%
  left_join(gdd.tree.smooth.80, 
            by = c("gdd", "p")) %>%
  ggplot() +
  geom_ribbon(aes(x = gdd, ymin = q10, ymax = q90),
              alpha = 0.7,
              fill = "darkgray") +
  geom_ribbon(aes(x = gdd, ymin = q2.5, ymax = q97.5),
              alpha = 0.7,
              fill = "darkgray") +
  geom_line(aes(x = gdd, y = p),
            color = "black") +
  geom_point(aes(x = gdd, y = obs_p),
             data = obs, size = 3) +
  # geom_text(aes(x = gdd, y = obs_p - 0.05, label = n),
  #           data = obs, size = 3) +
  labs(y = "Probability of detection",
       x = "Growing degree days") +
  scale_y_continuous(limits = c(-0.1,1)) +
  scale_x_continuous(breaks = c(500, 1000, 1500, 
                                2000, 2500, 3000, 3500)) +
  theme_bw() +
  theme(text = element_text(size = 14))
)

# save the plot
ggsave("figures/p_vs_gdd_gam.jpg", 
       width = 6, height = 5,
       dpi = 400)  
```
# Calculate number of samples required to achieve 95% detection by GDD
```{r}
# load the model fit
gdd.tree <- readRDS("output/gdd.tree.gam.rds")

# get posterior distributions for p at various predicted GDD values
preds <- posterior_epred(gdd.tree,
                re.form = NA,
                newdata = data.frame(gdd = seq(min(eab.gamdat$gdd),
                                               max(eab.gamdat$gdd),
                                               length.out = 100),
                                     tree = NA))

# calculate the number of days required to reach 95% certainty
    # for each of the 8000 iterations in the posterior
    # the formula is derived from the cumulative probabilty formula:
        # 1 - (1-p)^n where n is the number of samples
numsamps_list <- lapply(1:8000, 
                        function(x){
                          log(20) / (-log(1 - preds[x, 1:100]))
                          })

# summarize the resulting 8000 iterated calculations into quantiles
    # i.e., point estimate (50%) and credible intervals of 80% and 95%
numsamps <- do.call(rbind, numsamps_list) %>%
  apply(., 2, FUN = function(x) quantile(x, c(0.025, 0.1, 
                                              .5, 0.9, 0.975)))

# collapse that into a data.frame and add GDD back in
numsamps_df <- data.frame(gdd = seq(min(eab.gamdat$gdd),
                                               max(eab.gamdat$gdd),
                                               length.out = 100)) %>%
  cbind(t(numsamps)) %>%
  rename(q2.5 = 2, q10 = 3, days = 4, q90 = 5, q97.5 = 6)

# remove large unnecessary object
rm(numsamps_list)

# make the plot
(
numsamps.plot <-
numsamps_df %>%
  ggplot() +
  geom_ribbon(aes(x = gdd, ymin = q10, ymax = q90),
              alpha = 0.7,
              fill = "darkgray") +
  geom_ribbon(aes(x = gdd, ymin = q2.5, ymax = q97.5),
              alpha = 0.7,
              fill = "darkgray") +
  geom_line(aes(x = gdd, y = days),
            color = "black") +
  labs(y = "Samples required to \ndetect with 95% certainty",
       x = "Growing degree days") +
  scale_y_log10(breaks = c(2,5,10,25, 50,100,250, 500, 1000)) +
  scale_x_continuous(breaks = c(500, 1000, 1500, 
                                2000, 2500, 3000, 3500)) +
  theme_bw() +
  theme(text = element_text(size = 14))
)

# save the plot
ggsave("figures/samples95pct_vs_gdd_gam.jpg", 
       width = 6, height = 5,
       dpi = 400) 
```
# Make a composite figure of the 2 plots above
```{r}
library(patchwork)
gdd.plot2 <- gdd.plot + labs(x = "")
gdd.plot2 / numsamps.plot

ggsave("figures/both_vs_gdd_gam2.jpg", 
       width = 6, height = 8,
       dpi = 400) 

```
# Create Fig. - phenology of conventional detections 
```{r}
library(readxl)
cp <- read_xlsx("data/EAB_Phenology_FINAL.xlsx") %>%
  rename(p = 7, GDD = 9)

cp %>%
  filter(State != "WV") %>%
ggplot() +
  geom_point(aes(x = 32+GDD*9/5, y = p, color = State)) +
  geom_smooth(aes(x = 32+GDD*9/5, y = p),
              method = "gam", se = F,
              color = "black") +
  scale_color_manual(values = wesanderson::wes_palette("Zissou1")) +
  scale_x_continuous(limits = c(366,3584)) +
  scale_y_continuous(limits = c(0,0.4)) +
  labs(x = "Growing degree days", 
       y = "Proportion of total \nEAB trap catch") +
  theme_bw() +
  theme(text = element_text(size = 12))

ggsave("figures/conventional_trap_phenology.png", 
       height = 4, width = 6, dpi = 400)

```


###########
###############
Code below is a parallel analysis treating "singleton" PCR hits as zeros
###############
###########


# SUPPLEMENTAL: Run GAMs for detections vs. growing degree days
(Treat singleton PCR hit as non-detections.)
https://fromthebottomoftheheap.net/2018/04/21/fitting-gams-with-brms/
```{r}
# run a model with only a random intercept for "tree"
tree2 <- brm(bf(pos2 ~ (1|tree)),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, 
          control = list(adapt_delta = 0.99),
          save_pars = save_pars(all = TRUE))
saveRDS(tree2, "output/tree.gam_sup.rds") # save the output
stancode(tree2) # peek at the underlying stan code

# run the same model but now with a smoothed effect of GDD
gdd.tree2 <- brm(bf(pos2 ~ s(gdd) + (1|tree)),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, 
          control = list(adapt_delta = 0.99))
saveRDS(gdd.tree2, "output/gdd.tree.gam_sup.rds") # save it
stancode(gdd.tree2) # peek at the underlying stan code

# run the same model but now with a categorical effect of "direction"
gdd.tree.dir <- brm(bf(pos2 ~ s(gdd) + (1|tree) + dir),
          data = eab.gamdat, family = bernoulli(), cores = 3, seed = 17,
          iter = 4000, warmup = 2000, 
          control = list(adapt_delta = 0.99))
saveRDS(gdd.tree.dir2, "output/gdd.tree.dir.gam_sup.rds") # save it
stancode(gdd.tree.dir2) # peek at the underlying stan code

# read in the saved models (so we don't have to fit them each time)
tree2 <- readRDS("output/tree.gam_sup.rds")
gdd.tree2 <- readRDS("output/gdd.tree.gam_sup.rds")
gdd.tree.dir2 <- readRDS("output/gdd.tree.dir.gam_sup.rds")

# examine the output summary for each
tree2
gdd.tree2
gdd.tree.dir2

# compare the 3 models based on LOOIC
tree2 <- add_criterion(tree2, criterion = "loo",
                          moment_match = TRUE)
gdd.tree2 <- add_criterion(gdd.tree2, criterion = "loo",
                          moment_match = TRUE)
gdd.tree.dir2 <- add_criterion(gdd.tree.dir2, criterion = "loo",
                          moment_match = TRUE)

loo_compare(tree2, gdd.tree2, gdd.tree.dir2, 
            criterion = "loo") %>% 
  print(simplify = F)

# elpd_diff gdd.tree vs. gdd.tree.dir was xx (SE = xx)
#   vs. tree it was xx (SE = xx)
```
# SUPPLEMENTAL: Plot detection probability vs. GDD smooth
(Treat singleton PCR hit as non-detections.)
```{r}
# load the model fit
gdd.tree <- readRDS("output/gdd.tree.gam_sup.rds")

# make quick plots to inspect
gdd.tree.smooth <- conditional_smooths(gdd.tree, prob = c(0.95))
gdd.tree.smooth.e <- conditional_effects(gdd.tree, prob = c(0.95))
plot(gdd.tree.smooth)
plot(gdd.tree.smooth.e)

# get predictions vs. GDD with 95 and 80% credible intervals
gdd.tree.smooth.95 <- conditional_effects(gdd.tree, 
                                          prob = c(0.95))$gdd %>%
  select(gdd, p = estimate__, q2.5 = lower__, q97.5 = upper__)

gdd.tree.smooth.80 <- conditional_effects(gdd.tree, 
                                          prob = c(0.80))$gdd %>%
  select(gdd, p = estimate__, q10 = lower__, q90 = upper__)

# get observed proportions on each sampling date
obs <- eab.gamdat %>%
  group_by(as.character(gdd)) %>%
  summarise(obs_p = mean(pos2),
            n = length(pos2),
            .groups = "drop") %>%
  rename(gdd = 1) %>%
  mutate(gdd = as.numeric(gdd)) %>%
  arrange(gdd) %>%
  mutate(se = sqrt((obs_p * (1-obs_p))/n))

# make the plot
(
gdd.plot <-
gdd.tree.smooth.95 %>%
  left_join(gdd.tree.smooth.80, 
            by = c("gdd", "p")) %>%
  ggplot() +
  geom_ribbon(aes(x = gdd, ymin = q10, ymax = q90),
              alpha = 0.7,
              fill = "darkgray") +
  geom_ribbon(aes(x = gdd, ymin = q2.5, ymax = q97.5),
              alpha = 0.7,
              fill = "darkgray") +
  geom_line(aes(x = gdd, y = p),
            color = "black") +
  geom_point(aes(x = gdd, y = obs_p),
             data = obs, size = 3) +
  labs(y = "Probability of detection",
       x = "Growing degree days") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = c(500, 1000, 1500, 
                                2000, 2500, 3000, 3500)) +
  theme_bw() +
  theme(text = element_text(size = 14))
)

# save the plot
ggsave("figures/p_vs_gdd_gam_sup.jpg",
       width = 6, height = 5,
       dpi = 400)
```
# SUPPLEMENTAL: Calculate number of samples required to achieve 95% detection by GDD
(Treat singleton PCR hit as non-detections.)
```{r}
# load the model fit
gdd.tree <- readRDS("output/gdd.tree.gam_sup.rds")

# get posterior distributions for p at various predicted GDD values
preds <- posterior_epred(gdd.tree,
                re.form = NA,
                newdata = data.frame(gdd = seq(min(eab.gamdat$gdd),
                                               max(eab.gamdat$gdd),
                                               length.out = 100),
                                     tree = NA))

# calculate the number of days required to reach 95% certainty
    # for each of the 8000 iterations in the posterior
    # the formula is derived from the cumulative probabilty formula:
        # 1 - (1-p)^n where n is the number of samples
numsamps_list <- lapply(1:8000, 
                        function(x){
                          log(20) / (-log(1 - preds[x, 1:100]))
                          })

# summarize the resulting 8000 iterated calculations into quantiles
    # i.e., point estimate (50%) and credible intervals of 80% and 95%
numsamps <- do.call(rbind, numsamps_list) %>%
  apply(., 2, FUN = function(x) quantile(x, c(0.025, 0.1, 
                                              .5, 0.9, 0.975)))

# collapse that into a data.frame and add GDD back in
numsamps_df <- data.frame(gdd = seq(min(eab.gamdat$gdd),
                                               max(eab.gamdat$gdd),
                                               length.out = 100)) %>%
  cbind(t(numsamps)) %>%
  rename(q2.5 = 2, q10 = 3, days = 4, q90 = 5, q97.5 = 6)

# remove large unnecessary object
rm(numsamps_list)

# make the plot
(
numsamps.plot <-
numsamps_df %>%
  ggplot() +
  geom_ribbon(aes(x = gdd, ymin = q10, ymax = q90),
              alpha = 0.7,
              fill = "darkgray") +
  geom_ribbon(aes(x = gdd, ymin = q2.5, ymax = q97.5),
              alpha = 0.7,
              fill = "darkgray") +
  geom_line(aes(x = gdd, y = days),
            color = "black") +
  labs(y = "Samples required to \ndetect with 95% certainty",
       x = "Growing degree days") +
  scale_y_log10(breaks = c(2,5,10,25, 50,100,250, 500, 1000,
                           2500, 5000, 10000, 25000)) +
  scale_x_continuous(breaks = c(500, 1000, 1500, 
                                2000, 2500, 3000, 3500)) +
  theme_bw() +
  theme(text = element_text(size = 14))
)

# save the plot
# ggsave("figures/samples95pct_vs_gdd_gam_sup.jpg", 
#        width = 6, height = 5,
#        dpi = 400) 
```
# SUPPLEMENTAL: Make a composite figure of the 2 plots above
(Treat singleton PCR hit as non-detections.)
```{r}
library(patchwork)
gdd.plot2 <- gdd.plot + labs(x = "")
gdd.plot2 / numsamps.plot

ggsave("figures/both_vs_gdd_gam2_sup.jpg", 
       width = 6, height = 8,
       dpi = 400) 

```